Cleaning and Processing

### 2024-03-15 code/eda-code/eda-v1.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

@@

```{r}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(ggplot2) #for plotting
library(hrbrthemes) #for nice themes
hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
#rm(list=ls()) #clear the environment
```

@@

We are saving the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep.

@@

```{r}
# Summary stats of continuous variables
summary_stats <- ML_summary %>%
  summarize(
    mean_age = mean(Age, na.rm = TRUE),
    sd_age = sd(Age, na.rm = TRUE),
    mean_weight = mean(Weightgrams, na.rm = TRUE),
    sd_weight = sd(Weightgrams, na.rm = TRUE),
    mean_length = mean(Lengthcentimeters, na.rm = TRUE),
    sd_length = sd(Lengthcentimeters, na.rm = TRUE)
  )

@@

#Save table
saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))
```

Now let's work with our categorical variables.

```{r}
# Summary stats of categorical variables

# Calculate frequencies of color category (branca, parda, preta)
color_freq <- ML_summary %>%
  filter(!is.na(Color)) %>% # Exclude rows with NA in color variable
  count(Color, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(color_freq)
```

For maternal race, we have a sample size of n = 2695. The majority of the sample is White (branca), with 42.45% of the sample. Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample. Let's now compare White patients to all patients of color by combining Black and mixed-race categories. I will create a dummy variable for color, called modified_color with two categories: AfroDescent (comprising parda and preta) and EuroDescent (comprising branca).

```{r}
# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
  ifelse(ML_summary$Color == "Preta", "AfroDescent",
  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))

# View the updated dataframe
print(head(ML_summary))

# Calculate frequencies of color category (branca, parda, preta)
color_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedColor)) %>% # Exclude rows with NA in color variable
  count(ModifiedColor, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(color_freq2)
```
We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).

Now lets get summary statistics of the other categorical variables, including `Status`, `Nationality`, `Birth`, `MaternalOutcome`, `FetalOutcome` (excluding abortion), and `Sex`. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. For the `Status` variable, we will combine "multigesta" and "trigesta" into "multigesta", since they mean the same thing (more than two pregnancies). For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". "Brazilian" will remain the same.

```{r}
# Calculate frequencies of Status category
status_freq <- ML_summary %>%
  filter(!is.na(Status)) %>% # Exclude rows with NA in status variable
  count(Status, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(status_freq)

# Create dummy variables for status
ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Trigesta" = "Multigesta")

# Calculate frequencies of ModifiedStatus category
status_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedStatus)) %>% # Exclude rows with NA in status variable
  count(ModifiedStatus, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(status_freq2)
```

```{r}
# Calculate frequencies of Nationality category
nationality_freq <- ML_summary %>%
  filter(!is.na(Nationality)) %>% # Exclude rows with NA in nationality variable
  count(Nationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(nationality_freq)

# Create dummy variables for status
ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")

# Calculate frequencies of ModifiedStatus category
nationality_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedNationality)) %>% # Exclude rows with NA in nationality variable
  count(ModifiedNationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(nationality_freq2)
```
```{r}
# Calculate frequencies of birth category with abortion
birth_freq <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(birth_freq)
```
We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries. Now let's take abortion out of this sample size to just look at labor and delivery outcomes.

```{r}
# Calculate frequencies of birth category without abortion
birth_freq2 <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  filter(Birth != "aborto") %>% # Exclude rows with abortion
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(birth_freq2)
```
We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries. Now let's look at maternal outcomes.

```{r}
# Calculate frequencies of MaternalOutcome category
maternal_freq <- ML_summary %>%
  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(maternal_freq)
```
We can see that of the 2829 women who visited the clinic, 23 died, although not necessarily in childbirth. I need to come back to maternal mortality later after I have figured out live births. The equation is MMR = MD/LB * 10,000. We can also look at stillbirth rate. The equation is SBR = SB/LB * 1000.

```{r}
# Calculate frequencies of FetalOutcome category excluding abortion
fetal_freq <- ML_summary %>%
  filter(!is.na(FetalOutcome)) %>% # Exclude rows with NA in FetalOutcome variable
  count(FetalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  filter(FetalOutcome != "aborto") %>% # Exclude rows with abortion
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(fetal_freq)

#MMR Maternal mortality ratio
maternal_deaths <-23
live_births <- 2440
MMR <- (maternal_deaths / live_births) * 10000

#Stillbirth rate SBR
still_births <- 226
total_births <- 2666
SBR <- (still_births / total_births) * 1000
```
The MMR is 94.3% in this specific clinic. The SBR is 84.8% in this specific clinic.

Finally, lets look at infant sex.

```{r}
# Calculate frequencies of Sex category
sex_freq <- ML_summary %>%
  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category

print(sex_freq)
```

I will need to come back and calculate sex ratio in the clinic, but I need to first figure out how to exclude stillbirths from the sample, since sex ratio is calculated male/female * 100 for live births only.

>>>>

@@

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(hrbrthemes) #for nice themes
hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
#rm(list=ls()) #clear the environment
```

Load the data. First we will load the ML_summary set for summary statistics.

@@

We are saving the results to the `results` folder. All final table/figures will have `_final` in the name. This is to make sure we don't overwrite anything we might want to keep. All supplemental table/figures will have a `suppl` in the name.

@@

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*Branca*), the reference group here; mixed-race (*Parda*); and Black (*Preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category. For the `Status` variable, we will combine "multigesta", "secundigesta", and "trigesta" into "multigravida", since they mean the same thing (two or more pregnancies) and change the language to English (from gesta to gravida). We will also combine `secundipara` and `multipara` into one category `multipara`, since this means two or more viable pregnancies. The words in Portuguese and English here are the same. For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

```{r}
# Create dummy variables
# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
ML_summary$ModifiedColor <- ifelse(ML_summary$Color == "Branca", "EuroDescent",
  ifelse(ML_summary$Color == "Preta", "AfroDescent",
  ifelse(ML_summary$Color == "Parda", "AfroDescent", NA)))

# Create dummy variables for status and change to English
ML_summary$ModifiedStatus <- recode_factor(ML_summary$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")

# Create dummy variables for nationality and change to English
ML_summary$ModifiedNationality <- recode_factor(ML_summary$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")

# View the updated dataframe
print(head(ML_summary))
```

Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

```{r}
# Change observations from Portuguese to English
ML_summary <- ML_summary %>%
  mutate(
    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )
```

I have now learned that `gtsummary` is a very easy way to create summary statistics (Table 1). In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.

```{r, output="html"}
# Create gtsummary table 
table1 <- 
  ML_summary %>%
  tbl_summary(
    include = c(Color, ModifiedColor, ModifiedStatus, Age, Nationality, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Weightgrams, Lengthcentimeters),
    missing = "no",
    label = list(ModifiedColor ~ "Ancestry", ModifiedStatus ~ "Parity or Gravidity", Age ~ "Maternal Age", ModifiedNationality ~ "Combined Nationality", Birth ~ "Birth Outcome", MaternalOutcome ~ "Maternal Outcome", FetalOutcome ~ "Fetal Outcome", Weightgrams ~ "Infant Birthweight (grams)", Lengthcentimeters ~ "Infant Birth Length (centimeters)")
  ) %>%
  add_n()
table1

#Save table
saveRDS(table1, file = here::here("results", "tables", "table1_final.rds"))
```

For maternal race, we have a sample size of n = 2695. The majority of the sample is White (branca), with 42.45% of the sample. Mixed-race (parda) is 29.2% of the sample, and Black (preta) is 28.32% of the sample. Let's now compare White patients to all patients of color by combining Black and mixed-race categories. I will create a dummy variable for color, called modified_color with two categories: Afro-Descent (comprising parda and preta) and Euro-Descent (comprising branca).

We can see that of the clinic patients, 1551 (57.6%) were of African descent (defined as preta or parda) and 1144 (42.4%) were of European descent (branca).

We can see that of a sample size of n = 2761, 88% of outcomes were natural deliveries.

Now let's take abortion out of this sample size to just look at labor and delivery outcomes.

```{r}
# Calculate frequencies of birth category without abortion
birth_freq2 <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  filter(Birth != "abortion") %>% # Exclude rows with abortion
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(birth_freq2)
```
We can see that of a new sample size of n = 2672, 91% of outcomes were natural deliveries.

Now let's look at maternal outcomes.

Now let's look at some common reproductive and perinatal health statistics: maternal mortality ratio (MMR), stillbirth rate (SBR), and sex ratio at birth (SRaB). The equation is MMR = MD/LB * 10,000. We can also look at stillbirth rate. The equation is SBR = SB/LB * 1000.
```{r}
#MMR Maternal mortality ratio
maternal_deaths <-23
live_births <- 2440
MMR <- (maternal_deaths / live_births) * 10000
MMR

#Stillbirth rate SBR
still_births <- 226
total_births <- 2666
SBR <- (still_births / total_births) * 1000
SBR

#Sex ratio at birth (live male births/live female births)*100
sex_ratio <- ML_summary %>%
  filter(!Birth %in% c("Abortion") & !FetalOutcome %in% c("Stillbirth or Neonatal Death")) %>% # Exclude rows with abortion and stillbirth
 filter(!is.na(Sex) & !is.na(Birth) & !is.na(FetalOutcome)) %>%
  group_by(Sex) %>%
  summarise(count = n()) %>%
  spread(key = Sex, value = count) %>%
  mutate(SexRatio = M / F) * 100
sex_ratio
```
The MMR is 94.3% in this specific clinic. The SBR is 84.8% in this specific clinic. The sex ratio is 1.2 or there were 120 male live births per 100 female live births at this specific clinic. This is much higher than the current range of between 103 and 107 male births per 100 female births.

Because I am working on a different subset of the data, `ML_linear` for the statistical analysis, I will also need to recategorize variables here.

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear_processed.rds")
#load data
ML_linear <- readRDS(data_location)
```

```{r}
# Create summary
summary = skimr::skim(ML_linear)
print(summary)

# save to file
lineartable_file = here("results", "tables", "lineartable.rds")
saveRDS(summary, file = lineartable_file)
```

```{r}
# Create dummy variables
# Combine "Preta" and "Parda" into a new category Afro or EuroDescent and create dummy variable
ML_linear$ModifiedColor <- ifelse(ML_linear$Color == "Branca", "EuroDescent",
  ifelse(ML_linear$Color == "Preta", "AfroDescent",
  ifelse(ML_linear$Color == "Parda", "AfroDescent", NA)))

# Or using recode factor from dplyr (easier)
# ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" # = "AfroDescent", "Branca" = "EuroDescent")

# Create dummy variables for status and change to English
ML_linear$ModifiedStatus <- recode_factor(ML_linear$Status, "Secundigesta" = "Multigravida", "Trigesta" = "Multigravida", "Multigesta" = "Multigravida", "Secundipara" = "Multipara", "Multipara" = "Multipara", "Primigesta" = "Primigravida")

# Create dummy variables for nationality and change to English
ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "LatinAmerican", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "LatinAmerican", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "MiddleEastern", "Uruguaya" = "LatinAmerican")

# View the updated dataframe
print(head(ML_linear))

# Change observations from Portuguese to English
ML_linear <- ML_linear %>%
  mutate(
    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )
```

Here is the previous code I worked on before learning about `gtsummary`.

```{r}
# Don't use this code in future
# Summary stats of continuous variables
summary_stats <- ML_summary %>%
  summarize(
    mean_age = mean(Age, na.rm = TRUE),
    sd_age = sd(Age, na.rm = TRUE),
    mean_weight = mean(Weightgrams, na.rm = TRUE),
    sd_weight = sd(Weightgrams, na.rm = TRUE),
    mean_length = mean(Lengthcentimeters, na.rm = TRUE),
    sd_length = sd(Lengthcentimeters, na.rm = TRUE)
  )

@@

#Save table
#saveRDS(summary_stats_table, file = here::here("results", "tables", "summary_stats_table.rds"))

# Summary stats of categorical variables
# Calculate frequencies of color category (branca, parda, preta)
color_freq <- ML_summary %>%
  filter(!is.na(Color)) %>% # Exclude rows with NA in color variable
  count(Color, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(color_freq)

# Calculate frequencies of color category (branca, parda, preta)
color_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedColor)) %>% # Exclude rows with NA in color variable
  count(ModifiedColor, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(color_freq2)

# Calculate frequencies of Status category
status_freq <- ML_summary %>%
  filter(!is.na(Status)) %>% # Exclude rows with NA in status variable
  count(Status, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(status_freq)

# Calculate frequencies of ModifiedStatus category
status_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedStatus)) %>% # Exclude rows with NA in status variable
  count(ModifiedStatus, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(status_freq2)

# Calculate frequencies of Nationality category
nationality_freq <- ML_summary %>%
  filter(!is.na(Nationality)) %>% # Exclude rows with NA in nationality variable
  count(Nationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(nationality_freq)

# Calculate frequencies of ModifiedNationality category
nationality_freq2 <- ML_summary %>%
  filter(!is.na(ModifiedNationality)) %>% # Exclude rows with NA in nationality variable
  count(ModifiedNationality, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(nationality_freq2)

# Calculate frequencies of MaternalOutcome category
maternal_freq <- ML_summary %>%
  filter(!is.na(MaternalOutcome)) %>% # Exclude rows with NA in MaternalOutcome variable
  count(MaternalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(maternal_freq)

# Calculate frequencies of Sex category
sex_freq <- ML_summary %>%
  filter(!is.na(Sex)) %>% # Exclude rows with NA in Sex variable
  count(Sex, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(sex_freq)

# Calculate frequencies of birth category with abortion
birth_freq <- ML_summary %>%
  filter(!is.na(Birth)) %>% # Exclude rows with NA in birth variable
  count(Birth, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(birth_freq)

# Calculate frequencies of FetalOutcome category excluding abortion
fetal_freq <- ML_summary %>%
  filter(!is.na(FetalOutcome)) %>% # Exclude rows with NA in FetalOutcome variable
  count(FetalOutcome, na.rm = TRUE) %>% # Calculate frequencies while excluding NAs
  filter(FetalOutcome != "abortion") %>% # Exclude rows with abortion
  mutate(Total_Sample_Size = sum(n)) %>% # Calculate total sample size 
  mutate(Percent = n/sum(n)*100) # Calculate percentage of each category
print(fetal_freq)
```

### 2024-03-15 code/processing-code/processingfile-v1.qmd
Project Part 3
This is the first commit of all changes from Project Part 3.

# Structure of Data
glimpse(ML_raw)
```
From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

# Full column specification
spec(codebook)

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
```
We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model.

The latter (linear regression) will exclude the variables `MaternalOutcome` and `Birth`.

@@

# Creating dataset for linear regression
ML_linear <- ML_3 %>% tidyr::drop_na()

# Check for no missing data in linear model
ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(ML_linear_summary)
```
Now, I can see that my `ML_linear` dataset has no missing data. It has 2032 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.

>>>>

# Structure of Data
glimpse(ML_raw)
```

From this initial read, we can see that the data have 2845 observations and 23 variables. The variables are characterized as character and numerical.

@@

@@

```{r}
#Summarize missing data using dplyr
missing_data_summary <- ML_1 %>% dplyr::summarise_all(funs(sum(is.na(.))))
```

We can see that `CivilStatus` has 2834 missing observations and `GestationalAge_Months` has 2789 missing observations. This is a lot. Since I am not interested in understanding the relationship between marital status and infant birth weight and since gestational age in months refers to spontaneous abortions that occurred in the clinic, we can also exclude both these variables, the latter because an abortion precludes the existence of a birth.

@@

There are still missing observations for all the remaining variables except `Date`. For now, I will keep these observations intact because I want the maximum number of observations for summary statistics of all variables. Although the sample size for these summary statistics will then differ from that of my linear regression model, I believe more observations at the exploratory stage provide a fuller picture of what type of patient went to the hospital during the time period under study. Here, I will create two datasets: one for summary statistics and one for the linear regression model. In addition, most contemporary studies exclude all birthweights less than 500 grams because they have a high chance of being stillborn. I do so the same here, even if they were marked as a live birth. I will also exclude all stillbirths or neonatal deaths as the analysis is interested in studying live births.

@@

# Creating dataset for linear regression
ML_linear <- ML_3 %>% tidyr::drop_na() %>%
  filter(Weightgrams >= 500, FetalOutcome != "morto") # Drop observations of < 500 grams and all non-live births

# Check for no missing data in linear model
ML_Linear_summary <- ML_linear %>% dplyr::summarise_all(funs(sum(is.na(.))))
View(ML_linear_summary)
```

Now, I can see that my `ML_linear` dataset has no missing data. It has 1944 observations and 11 variables. This might change as I begin analysis. For example, I might want to exclude infant birth length, which is included at the moment.
